---
title: 'Eye on the Storm: A Journey Through Storm Trajectories'
author: "Nicole Rodriguez"
date: '04/17/25'
output:
  html_document:
    code_folding: show
    fig_caption: true
    df_print: kable
    theme: lumen
    toc: true
    toc_float:
      collapsed: false
  pdf_document:
    toc: true
---

# Introduction

This project explores the paths and intensities of major storms in the Atlantic using a combination of static and interactive visualizations. By leveraging R packages like **mapview()**, dynamic maps were created to visualize storm tracks, colored by attributes like *year* and *wind speed*. The interactive plots not only allow users to explore storm paths geographically, but also identify individual storms by clicking on their tracks. 

The data comes from the **storms** dataset available in the R package **dplyr**, which provides detailed information on storm names, tracks, wind speeds, and more. This dataset is important because it captures some of the most impactful storms, including Maria, Irma, Katrina, and Sandy. The visualizations provide valuable insight into the potential effects of climate change on the frequency and strength of hurricanes, an increasingly critical area of research.

***

# Data
## Load Libraries and Packages
```{r, warning = FALSE, message = FALSE}
# Load any R Packages 
if (!require("mosaic"))
  install.packages("mosaic")
if (!require("tidyverse"))
  install.packages("tidyverse")
if (!require("ggplot2"))
  install.packages("ggplot2")
if (!require("dplyr"))
  install.packages("dplyr")
if (!require("sf"))
  install.packages("sf")
if (!require("mapview"))
  install.packages("mapview")
if (!require("leafpop"))
  install.packages("leafpop")

library(mosaic) # Stats analysis
library(tidyverse) # Data manipulation & visualization
library(ggplot2) # Data visualization
library(dplyr) # Data manipulation
library(sf) # Simple features in R
library(mapview) # Interactive view
library(leafpop) # Popups with tables
```

## Description of Data
The storms dataset, available in the `dplyr` package, includes 13 variables and 19,537 observations including the position and attributes of storms from 1975-2022. This dataset offers valuable records of storm behavior over time, making it ideal for exploring patterns in storm paths, intensity. and how they may be changing over time. This is an important topic in the context of climate change and preparations for disasters. 

> R Package: `dplyr`

> Data set: `storms`

## Load & Filter Data
```{r, eval=FALSE}
# Load the storms dataset
storm_names <- dplyr::storms

# Remove storms with names starting with "AL0" or "Al2"
storm_names <- filter(storm_names, !grepl('AL0*', name) & !grepl('Al2*', name))

# Select wind, status, and name columns
storm_names <- storm_names %>%
  select(wind, status, name)
```

From the 216 hurricanes in this data set, we will be analyzing the top 6 well-known and most catastrophic hurricanes. 

> The six hurricanes featured in this analysis are: 

- **Sandy** 
- **Irma**
- **Andrew**
- **Maria**
- **Harvey**
- **Katrina**

```{r, warning = FALSE, message = FALSE}
# Load Data
storms_data <- dplyr::storms

# Filter storms
storms_data <- filter(storms_data, name == "Sandy" | name == "Irma" | name == "Andrew" | name == "Maria" | name == "Harvey" | name == "Katrina")

# Filter the storms_data dataset to include only tropical depressions, tropical storms, and hurricanes
storms_data <- filter(storms_data, status == "tropical depression" | status == "tropical storm" | status == "hurricane")

# Summary stats of wind by storm name
favstats(wind ~ name, data = storms_data)
```

```{r, warning = FALSE, message = FALSE, echo=FALSE}
# Convert storms_data to an sf (simple features) object using longitude and latitude, set CRS to WGS84
storms_data.sf <- storms_data %>% 
    st_as_sf(coords = c("long", "lat"), crs = 4326) 

# Create a POSIXct datetime column from year, month, day, and hour
# Then remove the now redundant individual columns
storms_data.sf <- storms_data.sf %>% 
    mutate(time = as.POSIXct(paste(paste(year, month, day, sep = "-"), 
                                   paste(hour, ":00", sep = "")))) %>% 
    select(-month, -day, -hour)

# Extract unique storm status values for use in color mapping
cls <- storms_data.sf %>% pull(status)
cls <- factor(cls, levels = unique(cls))  # Keep the original factor order
col = sf.colors(length(levels(cls)))      # Assign a unique color to each storm status

# Nest data by storm name and year, for individual storm tracking
storms_data.nest <- storms_data.sf %>% 
  group_by(name, year) %>% 
  nest()

# Function to convert grouped storm points into a single LINESTRING (storm track)
to_line <- function(tr) st_cast(st_combine(tr), "LINESTRING") %>% 
  .[[1]] 

# Apply the to_line function to each nested storm group to get all storm tracks
tracks <- storms_data.nest %>% 
  pull(data) %>% 
  map(to_line) %>% 
  st_sfc(crs = 4326)

# Create a new spatial object with storm metadata and geometry set to storm tracks
storms_data.tr <- storms_data.nest %>% 
  select(-data) %>% 
  st_sf(geometry = tracks)

# Function to create segment-wise LINESTRINGs between consecutive storm points
to_lines <- function(tr) { 
    g = st_geometry(tr)
    hd = head(g, -1)
    tl = tail(g, -1)
    map2(hd, tl, function(x, y) st_combine(st_sfc(x, y, crs = 4326))) %>% 
        map(function(x) st_cast(x, "LINESTRING"))
}

# Apply to_lines to each storm, flatten the list, and combine into one geometry set
trs <- storms_data.nest %>% 
    pull(data) %>% 
    map(to_lines) %>% 
    unlist(recursive = FALSE) %>% 
    do.call(c, .)

# Function to remove the last point from each storm path (to match line segments)
fn = function(x) head(x, -1) %>% 
  as.data.frame() %>% 
  select(-geometry)

# Update nested data to remove the final row (so it matches number of line segments)
storms_data.nest <- storms_data.nest %>% 
  mutate(data = map(data, fn))

# Unnest the cleaned data and bind with the segment-wise geometry
storms_data.tr2 <- storms_data.nest %>% 
  unnest() %>% st_set_geometry(trs)

```

* * *

# Data Analysis
## Storm Tracks by Status
```{r}
# Base R Plot mapping storms by storm status
plot(storms_data.sf["wind"], 
     type = 'p', # Point plot
     cex = 1, # Point size
     pch = 16, # Point shape
     axes = TRUE, # Show axes
     graticule = TRUE, # Grid lines
     main = "Storm Tracks by Status",
     xlab = "Longitude", # X-axis label
     ylab = "Latitude" # Y-axis label
     ) 

```



The plot above is a base R plot mapping the storm tracks colored by storm status. The following observations can be made: 

- The colors shift from deep blue to pink/yellow, suggesting that storms gain or lose intensity as they move.

- Some storms intensify (get more yellow) as they move north or over warm water.

- Yellow and bright pink points (indicating winds above 100 knots) are concentrated in the Caribbean Sea, Gulf of Mexico, and the western Atlantic, areas known for powerful hurricanes.

- Blue dots reappear in northern parts of the tracks, showing storms often weaken as they move north—typical as they hit cooler waters or land.

## Storm Tracks by Storm Name

```{r}
# Storm tracks colored by storm name with ggplot()
storms_data.tr %>%
  ggplot(aes(color = name)) + 
  geom_sf() + 
  theme(legend.position = "right") + 
  labs(title = "Storm Tracks by Storm Name", x = "Longitude", y = "Latitude")
```

The plot above was created with ggplot() to map the storm tracks colored by storm name.

- Red: Storm Andrew
- Brown: Storm Harvey
- Green: Storm Irma
- Light Blue: Katrina
- Dark Blue: Maria
- Pink: Sandy

> **Irma** had one of the longest tracks, originating from the far east and moved steadily to the west, and then headed northwest. 

> Many of these storms originated between 10°N and 20°N latitude and east of 80°W longitude (a known zone for hurricane formation). 

## Storm Tracks by Year

```{r}
# Hurricane tracks colored by year with ggplot()
storms_data.tr %>%
  ggplot(aes(color = year)) + 
  geom_sf() + 
  theme(legend.position = "right") + # Theme 
  labs(title = "Storm Tracks by Year", x = "Longitude", y = "Latitude") # Labels
```


The plot above was created with ggplot() to map the storm tracks colored by year. 

> **Harvey** formed in *1981*

> **Andrew** formed in *1992*

> **Katrina** formed in *2005*

> **Maria** formed in *2005*

> **Sandy** formed in *2012*

> **Irma** formed in *2017*


### Mapview of Storm Tracks by Year

```{r}
# Storm tracks colored by year with mapview()
mapview(storms_data.tr, 
        zcol = "year", # Color tracks by name & year
        layer.name = "Year",
        popup = storms_data.tr$name, # Add hurricane name to popup
        legend = TRUE) # Add legend

```


The interactive plot above was created with mapview() to map the storm tracks colored by year. This map provides a great visual to see the location of the storms on an actual map (Select any line to see which storm it is). It's also important to note that most of these storms occurred in the early 2000's.

> **Maria** & **Irma** have a long path, reflecting a longer tracking period and distance

> **Sandy's** track is known for it's significant turn back toward land, a major contributing factor in its impact


## Storm Tracks by Wind

```{r}
# Hurricane tracks colored by wind with ggplot()
storms_data.tr2 %>%
  ggplot(aes(color = wind)) +  
  geom_sf() + 
  theme(legend.position = "right") +  # Theme
  labs(title = "Storm Tracks by Wind", x = "Longitude", y = "Latitude")  # Labels
```

The plot above was created with ggplot() to map the storm tracks colored by wind. The higher the wind speeds, the lighter the blue. Some of the strongest storms with wind speeds greater than 100 were around 25°N latitude and between 70°W to 80°W longitude.

> **Irma** had the *highest* wind speeds while **Sandy** had the *lowest* wind speeds


### Mapview of Storm Tracks by Wind

```{r}
# Map storm tracks colored by wind with mapview()
mapview(storms_data.tr2, 
        zcol = "wind", # Color tracks by wind speeds
        layer.name = "Wind Speed (knots)", # Rename legend
        popup = popupTable(storms_data.tr2, 
                           zcol = c("name", "year", "status", "category", "wind")), # Popup table with stats
        legend = TRUE) # Add legend
```


The interactive plot above was created with mapview() to map the storm tracks colored by wind.  

> **Irma** reached wind speeds of **155 mph**

> **Katrina** reached wind speeds of **150 mph**

> **Maria** reached wind speeds of **150 mph**

> **Andrew** reached wind speeds of **150 mph**

> **Harvey** reached wind speeds of **115 mph**

> **Sandy** reached wind speeds of **100 mph**

* * *

# Additional Analysis
## Hurricanes after 2004
```{r}
# Filter data to have time greater than January 1, 2004
storms_data.tr3 <- storms_data.tr2 %>%
  filter(time >= as.POSIXct("2004-01-01"))
```

```{r}
# Storm tracks after 2004 colored by wind with ggplot()
storms_data.tr3 %>%
  ggplot(aes(color = wind)) +  
  geom_sf() + 
  theme(legend.position = "right") + # Theme
  labs(title = "Storm Tracks of Hurricanes after 2004 by Wind", x = "Longitude", y = "Latitude") # Labels
```


The plot above was created with ggplot() to map the storm tracks colored by wind; only including data to have time greater than January 1, 2004. The only storms that occurred after 2004 were:

- Katrina
- Maria
- Sandy
- Irma


### Mapview of Hurricanes After 2004 
```{r}
# Storm tracks colored by wind with mapview()
mapview(storms_data.tr3,
        zcol = "wind",
        legend = TRUE, # Add legend
        layer.name = "Wind Speed (knots)", # Rename legend
        popup = popupTable(storms_data.tr3, 
                           zcol = c("name", "year", "status", "category", "wind")) # Popup table
        )

```


The interactive plot above was created with mapview() to map the storm tracks colored by wind; only including data to have time greater than January 1, 2004. We can see that Katrina, Irma, Sandy, and Maria are shown on this map since they formed after 2004. All these storms developed to become hurricanes, with **Irma** reaching the strongest wind speeds of 155 mph in 2017. This data is important to research because it can provide insight on how climate change may be influencing the formation of stronger hurricanes over the years. 
